\documentclass[11pt]{amsart}
\usepackage{geometry}
\geometry{letterpaper}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{epstopdf}
\usepackage[T1]{fontenc}
\title{CS241}
\author{Andrew Codispoti}

\begin{document}
\maketitle
\section{Binary and Hexadecimal numbers}
\begin{enumerate}
  \item bit -- binary digits 1 and 0 (all computer understands)
  \item byte -- 8 bits
  \item word
    \begin{enumerate}
      \item machine specific grouping of bits
      \item assume 32-bit architecture
      \item 1 word = 32 bits= 4 bytes
    \end{enumerate}

  \item nibble -- 4 bits half a byte
\end{enumerate}


\subsection{Given a byte(or word) in memory what does it mean?}
\par Could mean many things.
\begin{enumerate}
  \item A number (which number?)
\end{enumerate}
\subsection{How can we represent negative numbers?}
\par Simply use a sign bit with 0 for + and 1 for - (Sign-Magnitude
representation) but then you have two -1's and arithmetic is tricky
\subsubsection{Two's Complement notation}
\par Interpret the n-bit number as a an unsigned int. If first bit is 0 done else
subtract $2^n$
\par n bits- represent $-2^{n-1}$ \dots $2^{n-1}$ with left bit still giving
sign. arithmetic is clean, just mod $2^n$
\par We cant tell if a number is signed unsigned or two's complement and we
have to remember.
\par We don't even know if what it means:a number, a character, An
instruction (or part of one), Garbage
\subsection{Hexadecimal notation}
\begin{enumerate}
  \item base 16 0-9, A-F
  \item more compact than binary
  \item each hex digit = 4 bits (1 nibble)
  \item e.g. 1100 1001  = C9
  \item NOTATION: 0xC9
\end{enumerate}

\subsection{Mapping from binary to characters}
\subsubsection {ASCII}
\par Uses 7 bits
\par IBM implemented extended ascii to use all 8-bits, but they add some weird
characters i.e. frame like characters.Compatibility issues because no one
standard.
\par 11001001 is not 7 bit ascii, 01001001 decimal 73 is ASCII for I
\par other standards like EBCDIC
\section{Machine Language}
\par Computer programs operate on data and are data(occupy same space as data)
\subsection{Von Neumann architecture}
\par Programs reside in the same memory as data.
\par Programs can operate on other programs i.e OS
\subsection{Central Processing Unit}
\par see physical notes for diagram
\begin{enumerate}
  \item Control Unit
    \begin{enumerate}
      \item decodes instructions
      \item dispatches to other parts of the computer to carry out instructions
    \end{enumerate}
  \item Arithmetic logic unit: Does Math
\end{enumerate}
\subsection{Memory--Many Kinds (Ranked in speed order)}
\begin{enumerate}
  \item **CPU
  \item cache
  \item **main memory RAM
  \item disk memory
  \item network memory
\end{enumerate}
\subsection{Registers}
\par On the CPU, small amount of very fast memory called registers
\par MIPS 32 General purpose registers \$0 to \$31
\begin{enumerate}
  \item each holds 32 bits
  \item can only operate on data that is in regs.
  \item \$0 is always 0
  \item \$31 is special and \$30
  \item EX: add the contents of two registers and put the result in another
    register.
  \item 5 bits encode a register $2^5 = 32$
  \item 15 bits to encode registers, 17 bits to encode operation
\end{enumerate}
\subsection{RAM}
\begin{enumerate}
  \item large amount of memory away from cpu
  \item travels between the cpu and ram on the bus
  \item big array of n-bytes, n ~ $10^9$
  \item each cell has an address 0,\dots, n-1
  \item each 4-byte block of the form is a word(see diagram 2 in notes)
  \item word addresses are 0,4,8,c,10,14,18,1c
  \item RAM access much slower than reg access
\end{enumerate}
\subsection{Communicating with RAM}
\par two commands
\begin{enumerate}
  \item load
    \begin{enumerate}
      \item transfer a word from an address to a register. desired address
        goes into the memory address register(MAR), goes out on bus
      \item data at that location comes back on the bus, goes into the memory
        data register (MDR)
      \item value in MDR moved to destination register
    \end{enumerate}
  \item store: does the reverse of load
\end{enumerate}
\subsection{How does a computer know which words contain instructions and which
  contain data?}
\par It Doesn't
\subsection{How does it run}
\par Special register called pc(program counter) which stores the address of
the next instruction to execute
instruction to execute.
\par By convention, guarantee that some address(i.e. 0) contains code,
initialize pc to 0.
\par Computer then runs the fetch-execute cycle
\begin{verbatim}
  PC <- 0
  loop
  IR <- MEM[PC]
  PC <- 4
  decode and execute the instruction in IR
  end loop
\end{verbatim}
\par only program the machine really runs
\par \textbf{NOTE:}  PC holds the address of the next instruction while the
current instruction is executing.
\subsection{How does a program get executed}
\par Program called a loader that puts the program in memory and sets PC to teh
address of the first instruction in the program and sets PC to teh address of
the first instruction in the program
\subsection{What happens when a program ends?}
\par need to return control to the loader, set pc to the address of the next
instruction in the loader.\$31 will contain the right address.
\par need to set pc to \$31
\subsection{Example}
\par Example1: Add value in \$5 to the value in the \$7 sotre result in \$3 and
return\\
\begin{tabular}{|c c c c|}
  \hline
  location & binary & hex & meaning\\
  \hline
  00000000 & 0000 0000 1010 0111 0001 1000 0010 0000 & 00a71820 & add \$3, \$5,
  \$7 \\
  \hline
  00000004 & 0000 0011 1110 0000 0000 0000 0000 1000 & 03e00008 * jr \$31 \\
  \hline
\end{tabular}
\par Example 2 add 42 to 52, store in \$3, return
\par lis \$d  "load immediate and skip", treat the next word as an immediate
value and load it in to \$d then skip to the instruction\\
\begin{tabular}{|c c c c|}
  \hline
  location & binary & hex & meaning\\
  \hline
  00000000 & 0000 0000 0000 0000 0010 1000 0001 0100 & 00002814 & lis \$5\\
  \hline
  00000004 & 0000 0000 0000 0000 0000 0000 0010 1010 & 00000004 & .word 42 \\
  \hline
  00000008 & 0000 0000 0000 0011 1000 1000 0001 0100 & 00002814 & lis \$7\\
  \hline
  0000000c & 0000 0000 0000 0000 0000 0000 0011 0100 & 00000004 & .word 52 \\
  \hline
  00000004 & 0000 0011 1110 0000 0000 0000 0000 1000 & 03e00008 & jr \$31 \\
  \hline
\end{tabular}
\begin{verbatim}
  add \$3 \$5 \$7
\end{verbatim}
\subsection{assembly language}
\par replace tedious binary/hex encodings with easier to read mnuemonics
less chance of error, translation to binary can be automated (assembler), one
line of assembly = one machine instruction (word)
\begin{verbatim}
  lis $5; load imm and skip
  .word 42; not an instruction, is a directive that next word in binary should
  literally be 42
  list $7;
  .word 52
  add $3 $5 $7;destination reg first
  jr $31
\end{verbatim}
\subsubsection{EX3}
\par Compute the absolute value of \$1 store in \$1 and return
\begin{enumerate}
  \item some insturcitons modify PC =  "branches" and "jumps", i.e. jr
  \item beq: branch if 2 registers have equal contents, increment PC by a given
    number of words, can branch backwards.
  \item lone bne
  \item slt: "set less than"
    \begin{verbatim}
      slt $a, $b, $c
      $a = {1 : $b<$c>, 0}
    \end{verbatim}

\end{enumerate}

\subsubsection{RAM}
\par lw= load word from ram into reg
\begin{verbatim}
  lw $a, i($b) loads the word at MEM[$b + i]into $a
\end{verbatim}

\par sw = store word from regs into RAM
\begin{verbatim}
  sw $a , i($b) stores word in $a at mem[$b+1]
\end{verbatim}
\par see examples on paper.
\subsection{Multiplication}
\par mult =  multiply
\begin{verbatim}
  mult $a, $b ; Product of 2 32 bit numbers up to 64bitz, too big for register
\end{verbatim}
\par Concatenation of hi and lo registers is the entire product of
multiplication
\par mflo = move from lo
\begin{verbatim}
  mflo $a, ; $a <- lo
\end{verbatim}
\par for division lo stores quotient, and hi stores remainder
\subsection{revisit looping example}
\par have to keep track of offsets if instuctions are added or remove
\subsection{Assembler}
\par assembler allows labeled instrcutions
\begin{verbatim}
  foo: add $1, $2, $3
\end{verbatim}
\par assembler associates the name foo with the address of the instruction
\par assembler calculates the distance between the label and the program
counter, in words $\frac{top-PC}{4}$
\section{Procedures in MIPS}
\par 2 problems to solve
\begin{enumerate}
  \item call and return : transferring control into and out of the procedure (
    Procedures calling other Procedures)
  \item Registers: what if a proc overwrites my registers.
\end{enumerate}
\par We could : reserve some registers for f and some for main line so they
wont interfere. but when using recursion we run out of registers.
\par Instead guarantee that procs leave regs unchanged when done by storing in
RAM. Must stop processes from using the same ram becuase the same issue will
arise.
\par see diagram
\par Can allocate from one end of ram to the other for procs. Need to track what
ram is in use. Mips machine helps us \$30 initialized by the loader to just
past the last word of memory.
\par can use \$30 as a "bookmark" to separate used and unused RAM
\par diagram
\par RAM uses LIFO order \$30 is the stack pointer address at the top of the
stack
\subsection{Template for Procedures}
\begin{verbatim}
  f:  sw $2, -4($30) ;
  sw $3, -8($30); push registers f modifies on the stack
  lis $3
  .word 8   ; decrement 30
  sub $30, $30, $3
  ; body

  g:
  add $30, $30, $3 ; assuming $3 remains 8 increment 30
  lw $3, -8($30)
  lw $2, -4($30)

\end{verbatim}
\subsection{Call and return}
\begin{verbatim}
  main: ...
  lis $5
  .word f; address of line labelled f
  jr $5 ; jump to that line
  ;(HERE)
\end{verbatim}
\par Return: we need to set PC to the line after the jr(i.e. to HERE)
\par Solution: jalr
\subsubsection{jalr(Jump and link regiter)}
\par like jr, but sets \$31 to the address of the next instruction
\begin{verbatim}
  main: ...
  lis $5
  .word f; address of line labelled f
  jalr $5 ; jump to that line
\end{verbatim}
\par  Question: jalr overwrites \$31 so how do we get back to the loader, what if f calls
g
\par Answer: Save \$31 to the stack before the call and restore afterwards

\begin{verbatim}
  main:
  lis $5
  .word f
  sw $31, -4($30)
  lis $31
  .word 4
  sub $30, $30, $31
  jalr $5
  lis $31
  .word 4
  add $30, $30, $31
  lw $31, -4($30)
  jr $31
\end{verbatim}

\subsection{Parameters and Results}
\par generally use regs (document)
\par if too many, use stack
\begin{verbatim}
  ; sum 1ToN: computes 1 + ... + N
  ; Register
  ; $1 - working
  ; $2 - input (value of N)
  $ $3 - output; do not save this one
  sum1toN:
\end{verbatim}
\section{Recursion}
\par no extra machinery needed
\par if registers, parameters, stack managed properly, recursion will just work
\subsection{I/O}
\par output:  Use sw to store word in location 0xffff000c. the last byte in the
word will be printed
\begin{verbatim}
  lis $1
  .word 0xffff000c
  list $2
  .word 67
  sw $2, 0($1)
  lis $2
  .word 83
  sw $2, 0($1)
\end{verbatim}
\section{The Assembler}
\par Any translation involves two phases
\begin{enumerate}
  \item Analysis: Understand what is meant by the source string
  \item Synthesis: output equivalent target string
\end{enumerate}
\par starts with assembly file: stream of characters
\begin{enumerate}
  \item group characters into meaningful tokens: label, hex \#, reg \#, .word, etc
  \item group tokens into instructions if posssible
  \item if tokens do not form sensible instructions, output ERROR to stderr
  \item NOTE: There are many more wrong tokens than right ones. try to find right
    combos
\end{enumerate}
\subsection{Biggest problem with assembler}
\par how do we assemble
\begin{verbatim}
  beq $2, $0, abc
  abc:
\end{verbatim}
\par cant assemble the first beq because we dont know the value of abc
\subsection{Standard solution}
\par assemble in two passes
\par Pass 1: group tokens into instructions. record addresses of labelled
instructions into a symbol table(list of [label, address] pairs)
\par NOTE: a line of assembly can have more than one label. you can label the
word after the end of the program
\par Pass 2: translate each instruction into machine code. If an insturction
refers to a label, lookup the associated address in the symbol table
\par Your Assembler output the assembled mips code to stdout. output the symbol
table to stderr
\par marmoset only cares outupt ERROr
\subsection{Code}
\begin{verbatim}
  main: lis $2
  .word 13
  add $3, $3, $0
  top: add $3, $3, $2
  lis $1
  .word 1
  sub $2, $2, $1
  bne $2, $0, top
  jr $31
  end:
\end{verbatim}
\subsubsection{Pass 1}
\par group tokens into instructions. build symbol table
\begin {tabular}{|c c|}
  \hline
  main &  0\\
  top & c\\
  end  & 24\\
  \hline
\end{tabular}

\subsubsection{Pass 2}
\par translate each instruction
\begin{verbatim}
  lis $2  ---- 0x00001014
  .word 13 ---- 0x0000000d
  bne $2, $0, top -- lookup top in symbol table
  calculate (top-pc)/4 = -5 ==> 0x1440fffb
\end{verbatim}
\section{bit level operations}
\par to assemble bne \$2, \$0, top (where (top - pc)/4  = -5)
\begin{verbatim}
  opcode = 000101 = 5
  1st reg  = $2 = 00010
  2nd reg = $0 = 00000
  offset = -5
  |6 bits (opcode) | 5 bits(1st reg) | 5 bits(2nd reg) | 16 bits (offset)
\end{verbatim}
\par to put 000101 into the first 6 bits append 26 0's (left shift by 26 bits)
5 << 26\\

\par move 2 21 bits to the left\\ 2 << 21
\par move 0 16 bits to the left\\0 << 16
\subsection{Bitwise and/or}
\par normal AND/OR. \\and with 0 gives 0, with 1 gives other digit\\ or with 0
gives other digit, 1 gives 1
\par use bitwise and to turn bits off
\par use bitwise or to turn bits on
\par bitwise and with 0xffff, -5 \& 0xffff bitwise and -5
\par bitwise or the four pieces together \\
(5 << 26) | (2 << 21) | (0 << 16) | (-5 \& 0xffff)
\subsection{C++ stuff}
\par int converts to int ascii codeo
\par char outputs the actual value and the screen interprets
\begin{verbatim}
  unsigned int instr = 5 << 26 |...;
  char c = instr >> 24;
  cout << c
  c = instr >> 16;
  cout << c;
  c = instr >> 8;
  cout << c;
\end{verbatim}
\section{Loaders}
\par OS Code:
\begin{verbatim}
  repeat:
  p <- next program to run
  copy P into memory, starting at 0
  jalr $0
  beq $0, $0, repeat
\end{verbatim}
\section{OS}
\par Problem: os is a program - where does it sit in memory. other programs in
memory at the same time, all cant be at address 0.
\par labels may be resolved to the wrong addresses
\subsection{How do you fix it}
\par could pick different starting addresses for programs at assembly time
\par let the loader decide where to put the program, fix bad label references
\subsection{Loader's job}
\begin{enumerate}
  \item take a program P as input, find a location a in memory for P
  \item copy P into memory starting at a
  \item return a to the os
\end{enumerate}
\begin{verbatim}
  OS 2.0
  repeat: p <- next program
  $3 <- loader(P)
  jalr $3
  beq $0, $0, repeat
\end{verbatim}
\section{Loader Pseudocode}
\par Input: words $w_1 , \dots, w_n$ < - the code
\par n = k + space for stack (how much? pick something)
\par a = address of n contiguous words if free RAM
\begin{verbatim}

  for i-0..k-1
  MEM[a+i*4] <- wi+1
  $30 a+4*n
  return a

\end{verbatim}
\subsection{What needs to change when we relocate?}
\begin{enumerate}
  \item offset added in order to  fix the askew entries. add alpha to word
  \item dont adjust constant word values
  \item do not adjust everythign else including branches
\end{enumerate}
\subsubsection{Problem}
\par Assmbled file is a stream of bits, how do you know it came from .word(with
an id!) and which are instructions
\par We cant do this, and we need more info from the assembler
\par output of most assemblers are not just machine code, it also produces
object code
Object file: contains binary code AND auxiliary info needed by the loader and
linker
\par we have our own object code format called MERL(Mips executable relocatable
linkable)
\subsubsection{What do we need to put in our object file}
\begin{enumerate}
  \item the code
  \item which lines of code(addresses) were originally .word id
\end{enumerate}
\section{MERL format}
\par start at 12 because header starts at 0 and it consists of 3 words
\subsection{Want assembler to generate relocatable object code}
\par Relocation Tool: cs241.merl ==> takes in MERL file and relocation address,
outputs non relocatable mips file with header and footer removed ready to load
at the given address
\par mips.twoints/array: optional second argument = address at which to load
the file
\par E.g.: load myobj.merl at 0x10000
\begin{verbatim}
  java cs241.merl 0x10000 < myobj.merl > myobj.mips
  java mips.twoints myobj.mips -x10000
\end{verbatim}
\par loader relocation algorithm
\begin{verbatim}
  read()//skip cookie
  endMod <- read() -12 //length of code and footer
  codeLen <- read() -12 //length of code
  alpha <- findfreeRAM(codeLen+stack)
  for(i=0; i<codeLen;i+=4>)
    MEM[a+i] = read()
  end for
  while(i<endMod)
  format <- read()
    if(format == 1)
      rel <- read()/address to be relocated
      MEM[rel+alpha - 12]/*actual location in RAM (header not
      loaded)*/+= alpha - 12    //adjust forward by alpha backward by header
    length 12
    else ERROR
    i+=8
  end while
\end{verbatim}
\section{Linkers}
\par Issue: how can the assembler resolve a reference to the label in a
different file
\par soln1: cat the files, assemble the result
\par soln 2: tool that understands Merl files and puts them together
intelligently - a linker
\par what should the assembler do with references to labels that aren't there?
\begin{enumerate}
\item need to change the assembler
\item when the assembler encounters .word id where label id is not found, it
  fills in 0 and indicates that the program requires the value of id before it
  can run
\end{enumerate}
\subsection{How does the assembler notify us?}
\par makes an entry in the MERL file
\par we lose a valuable error check by having to look at different files for
label definitions
\subsection{How can the assembler knwo what is an error and what is
  intentional?}
\par Create a new assembler directive:
\begin{verbatim}
  .import id
\end{verbatim}
\par tells the assembler to ask for id to be linked in.
\par Does NOT assemble to a work in MIPS
\par When the assemmbler encounters \verb|.word abc| if label abc: is not found
and no .import abc, then error. Notifyins us of imported symbols MERL entry
\par Format code 0x11 means External symbol reference(ESR)
\subsection{What information must be recorded?
}
\begin{enumerate}
  \item Where, at what address is the symbol being used? (where is the 0 that
    we need to fill in )
  \item what is the name of the symbol
\end{enumerate}
\par Format of an ESR entry:
\begin{verbatim}
  word1 - 0x11
  word2 location where the symbol is used
  word3 length of the name in characters
  word4
  .
  .       ASCII chars in the symbol's name (each char in its own word)
  .
  word 3+n
\end{verbatim}
\subsection{How can the linker know which abc to link to?}
\par labels wil sometimes be duplicated
\subsection{How can we make abc in b.asm unavailable}
\par another assembler directive and MERL entry type
\begin{verbatim}
  .export abc  - make abc available for linking
\end{verbatim}
\par does not assumble to a word of mips, tells te assembler to make an entry
in the MERL symbol table
\subsubsection{Entry Type: External Symbol Definition(ESD)}
\begin{verbatim}
  word1 - 0x05 - formate code for ESD
  word2 - address the symbol represents
  word3 - length of the name in chars(n)
  word4 -

  .     The name in ASCII, one word per char
  .
  word3+n
\end{verbatim}
\end{document}
